---
title: Benchmarking Loading HapMap-like Data into R
output:
pdf_document:
  toc: false
  highlight: zenburn
---

```{r}
library(rmarkdown)
library(microbenchmark)
library(ggplot2)
library(data.table)
```

## Create Test Data

For our benchmarking, we'll create a file that's in a format similar to the
HapMap format: a wide data format where many individuals' genotypes are
represented as columns. These data are mostly character (genotypes like A, T,
Y, or A/T), so in the tests we have the choice to read in as factors or
character vectors.

```{r}
nrows <- 1000L
ngeno <- 2500L

createFakeHapMap <- function(file, nrows, ngeno, nchr=12L, max_pos=1000000L) {
	pos <- data.frame(id=paste0("id_", seq_len(nrows)),
										chr=paste0("chr", sample(seq_len(nchr), nrows, replace=TRUE)),
										pos=as.integer(runif(nrows, 0, max_pos)))
	pos <- pos[order(pos$chr, pos$pos), ]
	geno <- do.call(cbind, lapply(seq_len(ngeno),
																function(x) {
																	rep("N", nrows)
																}))
	colnames(geno) <- paste0("ind_", seq_len(ngeno))
	out <- cbind(pos, geno)
	write.table(out, file, sep="\t", quote=FALSE,
							row.names=FALSE, col.names=TRUE)
}
file <- "fake-hapmap.txt"
gzfile <- paste0(file, ".gz")
createFakeHapMap(file, nrows, ngeno)
gzcmd <- sprintf("gzip -c %s > %s", file, gzfile)
system(gzcmd)

```

## Test Functions for Loading

We benchmark a few functions:

 - `readDelimFactor`: `read.delim` with `colClasses` and `nrows` set, reading
	 in genotypes as a factor.

 - `readDelimChar`: `read.delim` with `colClasses` and `nrows` set, reading in
	 genotypes as a character vector.

 - `readDelimNoRow`: `read.delim` with `colClasses` but not `nrows` set,
	 reading in genotypes as a character vector.

 - `readDelimNoColClasses`: `read.delim` without `colClasses` but with `nrows` set,
	 reading in genotypes as a character vector.

 - `readDelimNoRowNoColClasses`: `read.delim` without `colClasses` and `nrows` set,
	 reading in genotypes as a character vector.

 - `fread`: from the `data.table` package. This is very fast, but unfortunately
	 can't read gzipped files.

```{r}

HAPMAP_FIXED_COLS_CHAR  <- c(rs="character",
														 chrom="character", pos="integer")

HAPMAP_FIXED_COLS_FACTOR <- c(rs="character",
															chrom="factor", pos="integer")

readDelimFactor <- function(file, nlines) {
	header <- names(read.delim(file, header=TRUE, nrows=1))
	samples <- header[-c(1:length(HAPMAP_FIXED_COLS_FACTOR))]

	# create colclasses
	colclasses <- c(HAPMAP_FIXED_COLS_FACTOR, rep("factor", length(samples)))
	hapmap <- read.delim(file, header=TRUE, stringsAsFactors=FALSE,
											 colClasses=colclasses, nrows=nlines,
											 col.names=c(HAPMAP_FIXED_COLS_FACTOR, samples))
	return(hapmap)
}

readDelimChar <- function(file, nlines) {
	header <- names(read.delim(file, header=TRUE, nrows=1))
	samples <- header[-c(1:length(HAPMAP_FIXED_COLS_CHAR))]

	# create colclasses
	colclasses <- c(HAPMAP_FIXED_COLS_CHAR, rep("character", length(samples)))
	hapmap <- read.delim(file, header=TRUE, stringsAsFactors=FALSE,
											 colClasses=colclasses, nrows=nlines,
											 col.names=c(HAPMAP_FIXED_COLS_CHAR, samples))
	return(hapmap)
}

# don't use nrow=
readDelimNoRow <- function(file) {
	header <- names(read.delim(file, header=TRUE, nrows=1))
	samples <- header[-c(1:length(HAPMAP_FIXED_COLS_CHAR))]

	# create colclasses
	colclasses <- c(HAPMAP_FIXED_COLS_CHAR, rep("character", length(samples)))
	hapmap <- read.delim(file, header=TRUE, stringsAsFactors=FALSE,
											 colClasses=colclasses,
											 col.names=c(HAPMAP_FIXED_COLS_CHAR, samples))
	return(hapmap)
}

readDelimNoColClasses <- function(file, nlines) {
	header <- names(read.delim(file, header=TRUE, nrows=1))
	samples <- header[-c(1:length(HAPMAP_FIXED_COLS_CHAR))]

	# create colclasses
	colclasses <- c(HAPMAP_FIXED_COLS_CHAR, rep("character", length(samples)))
	hapmap <- read.delim(file, header=TRUE, stringsAsFactors=FALSE,
											 nrows=nlines,
											 col.names=c(HAPMAP_FIXED_COLS_CHAR, samples))
	return(hapmap)
}


readDelimNoRowNoColClasses <- function(file) {
	header <- names(read.delim(file, header=TRUE, nrows=1))
	samples <- header[-c(1:length(HAPMAP_FIXED_COLS_FACTOR))]

	# create colclasses
	hapmap <- read.delim(file, header=TRUE, stringsAsFactors=FALSE,
											 col.names=c(HAPMAP_FIXED_COLS_FACTOR, samples))
	return(hapmap)
}

```

## Benchmarks

```{r}
nlines <- nrows + 1L
file <- "fake-hapmap.txt"

res <- microbenchmark(
					readDelimFactor={o <- readDelimFactor(file, nlines)},
					readDelimFactorGzip={o <- readDelimFactor(gzfile, nlines)},
					readDelimChar={o <- readDelimChar(gzfile, nlines)},
					readDelimNoRow={o <- readDelimNoRow(file)},
					readDelimNoColClasses={o <- readDelimNoColClasses(file, nlines)},
					readDelimNoRowNoColClasses={o <- readDelimNoRowNoColClasses(file)},
					fread={o <- fread(file)},
					times=20L)

p <- ggplot(res) + geom_boxplot(aes(x=expr, y=time))
p + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


